---
title: "looping2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
#  <span style="color:blue"> A bit more on Looping 



# Loops can be "nested" on loop inside the other

Example: Calculate NPV for

* a range of different interest rates and 
* a range of damages 
* that may be incurred 10 years in the future

Steps

* define inputs (interest rates, damages)
* define output (NPV)
* write the function
* create a data structure to store results where we vary both interest rates and damages
* use nested for loops to fill in the data structure

Try it first...


```{r npvfor, echo=FALSE}

# write a function to compute npv
source("R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


#generate some input
damages = c(25,33,91,24)
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
yr=10

# compute some npv's for different discount rates
# first generate a dataframe to store results
npvs = as.data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# now use a for loop to populate
 for (i in 1:length(damages)) {
         for (j in 1:length(discount_rates)) {
       npvs[i,j]= compute_NPV(value=damages[i],       discount=discount_rates[j],time=yr )

         }
 }
 npvs
 
 
 #some data wrangling
colnames(npvs)=discount_rates
rownames(npvs)=damages
 npvs
 

 npvsg =npvs %>% pivot_longer(cols=everything(),names_to="dis", values_to="npv")
 head(npvsg)
 ggplot(npvsg, aes(x=npv, col=as.factor(dis)))+geom_density(size=2)+scale_color_brewer(type="seq", name="Discount")
 
 # how about summing all the damages
 npv.total =npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
 ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

 
 # we could also do this with sapply
 #generate some input
damages = c(25,33,91,24)
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
yr=10

# with purr family, map_dfc runs the function for each value and returns as a data frame 
npvs = damages %>% map_dfc(compute_NPV, time=10, discount=discount_rates)

# with apply (s apply is for a sequence)
alternate_npvs = sapply(damages, compute_NPV, time=10, discount=discount_rates)

 npvsg =npvs %>% pivot_longer(cols=everything(),names_to="dis", values_to="npv")
 head(npvsg)
 ggplot(npvsg, aes(x=npv, col=as.factor(dis)))+geom_density(size=2)+scale_color_brewer(type="seq", name="Discount")
 
 # how about summing all the damages
 npv.total =npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
 ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

 
 
```
