---
title: "Starting Functions: Class 3"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Some conventions (helpful later in the course)

* Always write your function in a text editor and then copy into R

* By convention we name files with functions in them by the name of the function.R
I called my function power_gen_orig so I'll save it to

e.g. **power_gen_orig.R**

* you can have R read a text file by source(“power_gen_orig.R”) - make sure you are in the right working directory

* keep organized by keeping all functions in a subdirectory called *R*

*  Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”


# Function Example

```{r  examples, echo=TRUE, include=TRUE}

# first set your working directory to just above the R directory
source("../R/power_gen_orig.R")

# to see the code
power_gen_orig

```

# Some insights into inputs

A cool thing is that you can use multiple values as inputs to your R function
Lets explore what it can do

```{r usingfunctions, echo=TRUE, include=TRUE}
# note that this works (calculates power for each height)
power_gen_orig(height=c(1,4,5), flow=2)

# and this works (calculates power for each flow rate)
power_gen_orig(height=4, flow=c(2,4,5))

# and this calculates power for each combination of flow and height
power_gen_orig(height=c(4,5,6), flow=c(2,4,6))

# but this doesn't work - why?
power_gen_orig(height=c(4,5,6, 8, 20), flow=c(2,4,6))

```

#
Note that by *sourcing* a function - it will essentially overwrite anything else in your workspace with the same name 

*Warning* - if you name your function, the same as an internal R function, your new function will take precidence, and **hide** the internal R function

Ideally choose function names that are likely to be unique 

If you overwrite a *base* package R function - good to delete it (see below)

You can also use *package_name*:: to get to a function that is hidden in the current workspace

#
```{r hidden, echo=TRUE, include=TRUE, error=TRUE}


# hiding the sort function

# normal use
mydata = c(1,5,8,22,2)
sort(mydata)

# define a new function that adds two numbers - silly - but lets call it sort
sort = function(a,b) {
  add = a+b
  return(add)
}

# now run
sort(mydata)

# how do we get the base package sort back
base::sort(mydata)

# or get rid of our silly function in the workspace
rm(sort)
sort(mydata)
```

# Documentation Best Practices

In additon to *in-line* documents - at the top of the * \*.R * where you define
you function, include information about the functions contract

* descriptions
* inputs and parameters
* outputs

With units and default values included

See example *R/power_gen_orig.R*

# <span style="color:orange">Documentation</span>

 
* There is a standard format for documentation that can be read by automatic programs (roxygen2) - an R package that generate “standard” R documentation - manual or help pages

* These automated approaches for building documentation (like roxygen2) and meta data (descriptions of data sets) are increasingly common - so you should get into the practice of being structured in your approach to documentation

* We will use the conventions that work with roxygen2 - and then use this program to generate formal R documentation. Roxygen is similar to Doxygen which is used for C code…so its a widely used format

Documentation is placed at the top of the *something.R* file
all lines start with *\#\'*

# <span style="color:orange">Documentation Components</span>

Later in the course we will make our own package and create help pages, you
can use the documentation at the top of your function definitions for this - so lets 
get into the *habit* now

Two parts

* **Description** - summary of what your model/function does
* **Tagged (using special “key” words)**

Here are some examples there are many others

* *\@param*  inputs/parameter description
* *\@return* what your function returns (outputs)
* *\@example*  how to use it
* *\@references*  citations or urls
* *\@author* YOU

(you don’t need all of these and there are more tags, but start with at least param and return, example is a good idea

See example

# <span style="color:orange"> Lets practice:: Automobile Power Generation </span>

Some background

One of the equations used to compute automobile fuel efficiency is as follows this is the power required to keep a car moving at a given speed is as follows:


$Pb = crolling * m *g *V + \frac{1}{2} * A*p_{air}*c_{drag}*V^3$


where crolling and cdrag are rolling and aerodynamic resistive coefficients, typical values are 0.015 and 0.3, respectively.

- V: is vehicle speed (assuming no headwind) in m/s (or mps) m: is vehicle mass in kg

- A is surface area of car $(m^2)$

- *g*: is acceleration due to gravity $(9.8 m/s^2 )$ 

- p_air = density of air $(1.2kg/m^3 )$

- Pb is power in Watts

What would be the function's contract - input/parameters? what will it do?

Write a function to compute power, given a truck of m=31752 kg (parameters for a heavy truck) for a range of different highway speeds (30 m/s, 80 km/hr, 120 km/hr, 40 km/hr, 5km/hr )
The surface area is 16 $m^2$

Plot power as a function of speed
Add a second line for lighter vehicle
Remember to keep the function in its own file, store it in a subdirectory called *R* and then create a Rmarkdown to use the function and store the Rmarkdown in a different directory

**Good Programming Practices**

Name variables so its easy to understand

Describe the contract at the top of the function (following the format for the *power_gen_orig.R* example

Also include the comments in the Rmarkdown that you are using to run the function

# <span style="color:orange"> PART 2: Review Two Rmarkdowns </span>

- Sampling_example.Rmd 
- Error_checking.Rmd
---
