---
title: 'Testing'
author: "Naomi Tague"
date: "January, 2020"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
#knitr::opts_knit$set(root.dir = '/Users/christina/Desktop/Courses.local/ESM262n/docs/packages')
library(tidyverse)
library(devtools)
library(roxygen2)
library(testthat)
install.packages("~/Courses/rainbow_0.1.0.tar.gz", repos = NULL, type = "source")

```



# <span style="color:orange">Testing</span>

Top 12 Reasons to Write Unit Tests - Burke and Coyner (Java programmers)



* Tests reduce bugs in new features
* Tests reduce bugs in existing features
* Tests defend agains other programmers
* Testing forces you to slow down and think
* Testing makes development faster
* Tests reduce fear

Model developers now often use software to help them automate the testing process
Re-uses tests - makes it efficient to repeat many tests as you develop and modify the code 

Particularly helpful when you have multiple modules (as in our mangrove example)

This type of software is available for R, Python, C etc.

# <span style="color:orange">Testing in R</span>

In R, *testthat* library 
install and add this library

If you are in the working directory where you’ve stored the files in a specific directory (e.g **R**) you can use *testthat* to automate testing of your functions

* need devtools and “testthat” libraries
* test_dir(“name”) :runs all tests in the “name” subdirectory  (all files beginning with the word “test”
* test_file(“name”): runs all  the tests in a file called “name”

Testing in R uses expectation

# <span style="color:orange">Expectation</span>


tests you can use to make sure your code is working the way you think it should be working

* expect_true(condition)
* expect_equal(value, value)
* expect_match(string1, string2)
* expect_null(value)
* expect_length(object, n)



```{r testingsimple, eval=TRUE, error=TRUE}


# runs a test on a statement and checks to see if it is correct
# this one should work
expect_equal(2+2, 4)
# this one should 'throw and error'
expect_equal(2+5, 4)

# we can also use expect_true
# works with variables two
a = 200
expect_true(a > 2 )
a = 0
expect_true(a > 2 )

# matches
fish1 = "salmon"
fish2 = "trout"
expect_match(fish1, fish2)

# length
a = seq(from=1, to=10)
expect_length(a, 10)
expect_length(a, 11)

```

# The art of testing

It is good practice to create tests for functions (especially
useful if you are going to have multiple functions and/or multiple developers)

It make sure that the function returns what you think

Recall when we designed contracts at the beginning of class, testing is a way to make sure the program is honoring the contract

You need to be creative here - test for things you know will work (quick way to find bugs)

Usually we want to run multiple tests..

# <span style="color:orange">Testing in R</span>


A **test** runs an **expectation** function and returns a message

Format

test_that(description, series of expectations)
```{r testing2, eval=FALSE}


source("R/compute_NPV.R")
# lets do a simple test if profit is 0 , NPV should be 0
# if time is 0, NPV should be the original value
test_that("compute_NPV_works", {

expect_equal(compute_NPV(value=0,time=100,discount=0.1), 0)
expect_equal(compute_NPV(value=100, time=0, discount=0.1), 100)
})



```
# <span style="color:orange">Using tests in package development</span>

What we really want is to have tests that we **always** run, everytime we change something


Store multiple test in a *name.R* file  store these in these in the *tests/testthat* directory

Two options

To run all tests in a file, use *test_file*


To run all tests in a directory,use *test_dir*  (you can also do this from Build menu)



Goal: Make tests for each function in your package

```{r testing, eval=FALSE}
clim_data = as.data.frame(cbind(month=c(1:4), day=rep(1, times=4), year=c(2000,2000,2001,2001), 
                                 precip=rep(0, times=4), tavg=c(2,2,2,4)))
spring_summary(clim_data)
spring_summary(clim_data, spring_months=c(1,4))

#save this to a file called test_spring_summary
test_that("spring_summary_works" ,
{clim_data = as.data.frame(cbind(month=c(1:4), day=rep(1, times=4), year=c(2000,2000,2001,2001), 
                                 precip=rep(0, times=4), tavg=c(2,2,2,4)))

expect_equal(spring_summary(clim_data, spring_months=4)$Pavg_spring,0)
expect_equal(spring_summary(clim_data, spring_months=4)$Tavg_spring, 4)
expect_equal(spring_summary(clim_data, spring_months=2)$Tavg_spring, 2)
expect_equal(spring_summary(clim_data, spring_months=c(1:4))$warmest_spring, 2001)

})

# see example with rainbow package
# this part can only be done if we are in build model with the rainbow package
# to run the tests
#test_file("./Tests/testthat/test_spring_summary.R")

# do all tests in a directory
#test_dir("Tests/testthat", reporter="tap")

# you can also run Test under build menu
