
---
title: 'Programming Basics and Best Practices'
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
  html_document:
    df_print: paged
---

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# note start with index.Rmd
```




# Protocols and Guidance on how to best use this class

*  Learning to program is hard and I may not always explain in a way that is accessible to you - So if you don't understand something *ASK* , you will help me to learn how to explain (or find an answer if I don't know it) and you will help others

* Programming gets better with practice and play - Don't just read the Rmarkdown - try the code, try variations on the ideas presented, make up stuff to try, get your feet wet 

* Programming means making mistakes, expect it, stay calm and try again - if you get frustrated step away and come back; be creative


* Support each other 
  * when working in groups, pay attention to your partner, if they are not at your skill level, help them learn - recognize that we all have different backgrounds
  

* If you are really struggling, reach out to Ojas or myself, we can help
(or if you just want to chat about something ) 




# Why learn to code

* some evidence that computational thinking enhances
  * problem solving
  * critical thinking
  * collaboration
  * project management

[Learning to code or coding to learn? A systematic review, Popat and Starkey, 2019](https://www.sciencedirect.com/science/article/pii/S0360131518302768?casa_token=627kIJVKiUkAAAAA:QJvPu4by9dRLrueB8yLf769wWcnrJFwdn2IzZI5pzCViFRjzG5Ilj5tFSEQZ4ujDWnc2VGs)


# What is a program

* A program is a set of instructions for a computer to follow
* Programs are often used to manipulate data (in all type and formats you discussed last week)
* Simple to complex

    + scripts that you save in R-Markdown
    + instructions to analyze relationships in census data and visualize them
    + a model of global climate
 
* A key skill is being able to break tasks down into tasks (this also makes using AI-tools much easier **prompt engineering**

# Programming Languages

* Programs can be written in many different languages (all have their strengths and weakness)
* Languages expect instructions in a particular form (syntax) and then translate them to be readable by the computer
* Languages have evolved to make it help users write programs that are easy to understand, re-use, extend, test, run quickly, use lots of data...

# Programming Fundamentals - "Big Ideas"

* Data structures (simple variables, arrays, lists...) - how do we organize data
*  Basic Operations (=,+,-,...concatenate, copy)  to Control structures (if then, loops) - how do we manipulate data
* Modules...Functions - how do we make all this efficient - especially when we want to similar types of things over an over

These concepts common to all languages through the syntax may be different




# ChatGPT and other AI-tools



[Interesting Paper on AI-tools in Computing Education](https://dl.acm.org/doi/pdf/10.1145/3624720)

Tools 

* Github Copilot (most widely used by programmers)
* ChatGPT

(both built from CODEX by OpenAI)

In 2020, test CODEX scored 78% on intro to computing test, in 2023 it scored 98%

Other studies show less accuracy...depends on the problem

# Learning to code/programming helps you think about data science

"Metacognition, or “thinking
about thinking”, is a key aspect of
computational thinking (and problem-solving in general)"

"Relying
too heavily on code-generation tools
may hinder the development of these
crucial metacognitive skills" (Denny et al, 2024)

# So ...what should your strategy be

* AI-tools great resources but effective use requires understanding computing/problem solving

* Use this course to build that understanding, think about the meta concepts. Ask yourself - What are situations you might apply a technique/idea/concept;

* You gain skills by <span style="color:red">  Practicing 

* For assignments/practice
  - **Always** write the code yourself -  but you use AI-tools to help find syntax
  - Use AI-tools to explain other peoples code (including class examples)
  - Use AI-tools to help interpret error messages



  
# Best Practices for Software Development 


* Wilson G, Aruliah DA, Brown CT, Chue Hong NP, Davis M, et al. (2014) Best Practices for Scientific Computing. PLoS Biol 12(1): e1001745. doi:10.1371/journal.pbio.1001745

*  Blanton, B and Lenhardt, C 2014. A Scientist’s Perspective on Sustainable Scientific Software. Journal of Open Research Software 2(1):e17, DOI: http://dx.doi.org/10.5334/jors.ba

[Alternative View](http://simpleprogrammer.com/2013/02/17/principles-are-timeless-best-practices-are-fads/)

# Summary of Best Practices

[Wilson et al, Summary Box](https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.1001745#pbio-1001745-box001)
  

# Common Coding/Software Engineering Issues

* Unreadable code (hard to understand, easy to forget how it works, hard to find errors, hard to expand)
* Overly complex, disorganized code (hard to find errors; hard to modify-expand)
* Insufficient testing (both during development and after)
* Not tracking code changes (multiple versions, which is correct?)


# Automated Tools for Best Practices in Software Engineering

Automated tools (really useful for more complex code development
( note that GP’s often create programs > 100 lines of code)

Automated documentation

http://www.stack.nl/~dimitri/doxygen/

http://roxygen.org/roxygen2-manual.pdf 

Automated test case development

http://r-pkgs.had.co.nz/tests.html


Automated code evolution tracking (Version Control)
https://github.com/



# Inclass task

Download and read in "sagedata.txt" from Canvas
In an Rmarkdown document, do something with this data
Whatever is interesting for you

Try to do something that is at least a two-step process (e.g manipulating the data in some way then graphing)

**Add comments to your Rmarkdown to explain what you are doing, why - so that someone else could follow your code **

# The data

**p301.csv**

Climate and estimates of water and forest carbon information from USFS experimental watershed  in the Southern California Sierra (above Fresno )

Upload your Rmarkdown to our class shared drive
name it "yourname.Rmd"


# What the variables are (metadata)

* month

* year

* day

* wy  (Water Year - a hydrologic way of counting years that goes from Oct-Sept (instead of Jan-Dec))

* tmax  Maximum daily Temperature in degrees C

* tmin  Minimum daily Temperature in degrees C

* precip Depth of precipitation in mm

* transpiration  Estimated forest transpiration (m/m2/day)
* evaporation  Estimated evaporation (e.g ET would be transpiration + evaporation) (m/m2/day)

* NPP Estimated net carbon uptake by trees (kgC/m2/day)

* plant_carbon Estimate standing biomass in units of carbon (kg/m2)

* streamflow (m/m2/day) Streamflow (per watershed area)

* snowpack (m/m2) Amount of snow on the ground (in units of water)

* direct solar (kJ/m2/day) Incoming direct solar radiation
* diffuse solar (kJ/m2/day) Incoming diffuse solar radiation

Pair and share - 

# Here's my example 

a) providing detailed comments 
b) how I use ChatGPT

```{r, echo=TRUE}

library(tidyverse)
# read in the data - call it something meaningful
provwater = read.csv("../data/p301,csv", header=T)

# always check to make sure you read it in correctly
# looking at header, first and last lines can help
names(provwater)
#head(provwater)
#tail(provwater)
```


Since plant require sunlight, lets see how NPP varies with solar radiation (will focus on direct first) - when looking at relationships between variables, simply plotting as points with x-y type graphs is helpful (dependent variable on y-axis - here NPP is dependent on solar radiation)
```{r, echo=T}

ggplot(provwater, aes(direct_solar, NPP))+geom_point()

# make the graph prettier/more informative by adding labels
# also add a trend line since it looks like there is a relationship - non-liner 

# Ask ChatGPT - how I add a non-linear best fit line to a ggplot graph
ggplot(provwater, aes(direct_solar, NPP))+geom_point(size=2, col="green")+labs(x="Incoming direct radiation (kJ/m2/day", y="NPP (plant caron flux) kgC/m2/day")+geom_smooth(method="gam")

# lets look at how minimum daily temperature is changing through time
ggplot(provwater, aes(wy, tmin))+geom_point()

#kind of messy, because of day to day variation; maybe hides long term trends

# lets aggregate to annual; while I'm doing it I'm adding some other variables in case I want to look at them later

# ask ChatGPT - how do I average variables by  water year a data frame that includes water year as a variable
provwater_wy = provwater %>% group_by(wy) %>% summarize(tmax=mean(tmax), tmin=mean(tmin), precip=sum(precip), streamflow = sum(streamflow), peakSWE=max(snowpack))

# plot summarizes through time to look for trends
ggplot(provwater_wy,aes(wy, tmin), col="blue")+geom_point()

# maybe there is a trend (e.g linear trend in tmin with water year)
# test if its statistically significant

# ask ChatGPT
# what is the format in R for using a linear model to test if a slope is significant 
summary(lm(data=provwater_wy, tmin~wy))

```
