---
title: 'Using functions with Data'
author: "Naomi Tague"
date: "January, 2020"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```


# Comments/Questions 



# Review

* We are using functions as a way to design modular programs (that might involve many functions that are linked together)

* We always start with clearly defininng our goal for the whole program and each piece (and diagramming this if needed for programs made of multiple functions)

* Always write your function in a text editor and then copy into R

* By convention we name files with functions in them by the name of the function.R
e.g. **power_gen_orig.R**

* you can have R read a text file by source(“power_gen.R”) - make sure you are in the right working directory

* keep organized by keeping all functions in a subdirectory called *R*

*  Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”




```{r}

source("R/power_gen_orig.R")
```


# Multiple inputs 

For a function with more than one input  R can handle multiple values


**IF** 

* all (multiple) inputs have the same number of values 
* all other inputs or parameters have a single value


For other cases, additional instructions would need to be included in the function to help R understand what to do


Consider our reservoir function  ( **R/power_gen_orig.R**)

```{r sampling2}

#source("R/power_gen_orig.R")

power_gen_orig= function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  
  # calculate power
  result = rho * height * flow * g * Keff
  return(result)
}

# single
power_gen_orig(height=2, flow=2)

# multiple heights
# generate a structure to store results
# consider all heights from 1 to 100

resv_results = data.frame(height=seq(from=1, to=100))

# compute power generated for EACH height
resv_results$power = power_gen_orig(resv_results$height, flow=2)

# plot
ggplot(resv_results, aes(height, power))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")

# random combinations of height and flow
# use rnorm to sample from a normal distribution
resv_results$height = rnorm(mean=10, sd=2, n=100)
resv_results$flow = rnorm(mean=20, sd=1, n=100)
# compute power generated for EACH height
resv_results$power = power_gen_orig(resv_results$height, flow=resv_results$flow)

# plot
ggplot(resv_results, aes(height, power, col=flow))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")

# Unequal lengths - this will "work" but inputs get recycled unclear what is happening
# NOT recommended

eheight = seq(from=1, to=50)
eflow = runif(min=0.1, max=2, n=100)
resv_result = power_gen_orig(eheight, eflow)

# better
eheight = seq(from=1, to=50)
eflow = runif(min=0.1, max=2, n=50)
resv_result = power_gen_orig(eheight, eflow)

# another example of the problem of unequal length inputs
power_gen_orig(eheight, eflow, g=c(9,10,11))

#better
g = runif(min=8, max=10, n=50)
resv_result = power_gen_orig(eheight, eflow, g)
ggplot(resv_results, aes(height, power, col=flow))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")


```

# Error Checking

What will your function do if user gives you garbage data, imagine that you have a function that computes microbial growth rates with temperature

* Two options

<span style="color:skyblue">
1. error-checking (nice)
</span>

 if temperature < -100 or > 100
 
*  set results to NA
*  or exit function with a warming
 
<span style="color:skyblue"> 
2. assume user reads the contract :) return unrealistic values
</span>

if input -999.0, will still try to output growth rate
 
 
 Error-checking is helpful if you are going to build a model made up of many functions- why?
 

# How to add error checking

Consider this example (I'm going to put the function definition inline for now so you can easily see it - later we will save to our R subdirectory for reuse)

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   if (height < 0) height=NA
  
 # an alterative is to exit the function prematurely 
  if (flow < 0) return("Flow cannot be less than 0")
  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)
```


While this works for a case with one height and one flow
It will have problems if we use multiple heights - (a later topic)

A better implementation if you think the function might be run for multiple heights - YES you can do that

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   height = ifelse( (height < 0), NA, height)
  
 # an alterative is to exit the function prematurely 
  flow = ifelse((flow < 0), return("flow must be greater than zero"), flow)

  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)


# note that this works (calculates power for each height)
power_gen(height=c(1,4,5), flow=2)

# and this works (calculates power for each flow rate)
power_gen(height=4, flow=c(2,4,5))

# and this calculates power for each combination of flow and height
power_gen(height=c(4,5,6), flow=c(2,4,6))

# but this doesn't work - why?
power_gen(height=c(4,5,6, 8, 9, 20), flow=c(2,4,6))

# we could add error checking to test for this

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  
  # make sure we have same number of flow and height's so we can match them
  if (length(height) != length(flow)) return(" number of heights is different from number of flows")
  # make sure inputs are positive and if not set as NA so result will be NA
   height = ifelse( (height < 0), NA, height)
  
 # an alterative is to exit the function prematurely 
  flow = ifelse((flow < 0), return("flow must be greater than zero"), flow)

  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(height=c(4,5,6, 8, 9, 20), flow=c(2,4,6))
power_gen(height=c(4,5,6), flow=c(2,4,6))
```

See **power_gen.R** for file with error checking

Error checking is good practice - and combinded with good documentation - makes programming much much easier!





# Steps for running your model over multiple inputs

1. design a data structure to store results: sometimes this is automatic but not always
2. generate the input data
3. apply to the model



# Consider our function that estimates the power required to keep a car moving at a given speed



# Code

```{r powerexample}

#' Power Required by Speed
#'
#' This function determines the power required to keep a vehicle moving at 
#' a given speed
#' @param cdrag coefficient due to drag default=0.3 
#' @param crolling coefficient due to rolling/friction default=0.015
#' @param v vehicle speed (m/2)
#' @param m vehicle mass (kg)
#' @param A area of front of vehicle (m2)
#' @param g acceleration due to gravity (m/s) default=9.8
#' @param pair (kg/m3) default =1.2
#' @return power (W)

autopower = function(V, m, A, cdrag=0.3, crolling=0.015,pair=1.2,g=9.8) {
  P = crolling*m*g*V + 1/2*A*pair*cdrag*V**3
  return(P)
}

```

I've also stored the autopower.R function as a file = using the recommended naming convention

**autopower.R**


# Generating data for the function and iterating over that data

Two parts

* Generate a data structure to store your results

* Repeat application of your function over the data


Example use: Imagine we want to see how much power is generated given scenario where we know the mean and standard deviation of vehicles speeds


```{r sampling}

source("../R/autopower.R")

# generate sample speeds from a distribution
nsample = 100
speeds = rnorm(mean=25, sd=4, n=nsample)

# Step 1  create data frame to store results 
# how many simulations, what do you want to keep

#create a dataframe that has rows for each model run
# columns for height, flowrate and power estimate
results = data.frame(speed=speeds, power=NA)

head(results)

# if you only have one input and everything else is the same, R is smart enough
# to create a set of outputs
results$power = autopower(V=speeds, A=25, m=20000)

# ALWAYS look at your results to make sure it make sense
ggplot(results, aes(speed, power/1000))+geom_point()+labs(x="speed in m/s", y="power in kW")
ggplot(results, aes(x="Across Speed",y=power/1000))+geom_boxplot()+labs(y="power in kW")

```

# Looping (or repetition) 

Looping - or repeating applying your function over multiple inputs is not always automatic 
  * more complex scenarios - number of different values is not the same for all inputs
    (e.g - you have 3 different speeds for 5 different cars (mass/Area))
  * other programming languages

Lets start with the basics of how you do repetition in programming in general
often called looping

```{r loop }

# repeat statement
a=0
for (i in 1:5) {
 a = a+i
}
a

#if you want to keep track of a for each iteration
# start with a data structure to hold the results - there will be 5 iterations

a = rep(x=0, times=5)
for (i in 1:5) {
 a[i] = a[i]+i
}
a

# lets say we wanted to look income minus expenses for net income
# create some example

income = runif(min=1000, max=5000, n=10)
expenses = rnorm(mean=1500, sd=500, n=10)
# it is possible that this approach could generate negative expenses
expenses = ifelse(expenses < 0, 0, expenses)
# normally we can just do
net = income-expenses
# lets make this a data frame for nice plotting
account = data.frame(income=income, expenses=expenses, net = net)
ggplot(account, aes(income, expenses))+geom_point()
# lets add a year
account$year = seq(from=2000, length.out=nrow(account))
# we could have computed net with a for loop
# sometimes we use NA to show we haven't computed it yet, so "initialize" as NA
netloop = rep(NA, times=10)
# alternatively we could add to account
account$netloop=NA
account
# note I can use any variable I want to as a counter - notice how I'm now using columns of the 
# data frame
for (n in 1:10) {
  account$netloop[n] = account$income[n]-account$expenses[n]
}

# as expected net and netloop are the same :)
head(account)

# why loop
# what if we wanted to combine different rows in some way - 
# for example, we wanted to carry forward from the previous year

# so carryforward is current years net  plus last years net

# note: we don't have to do the first year as we don't have carry forward
# initialize our column to zero, start counter (n) at 2 instead of 1
# make the first value equal to the net of first year
account$carryforwardnet = 0
account$carryforwardnet[1]=account$net[1]
# now loop - looking backward one year to get previous years carry forward
for (n in 2:10) {
  account$carryforwardnet[n] = account$net[n]+account$carryforwardnet[n-1]
}

# plot - and use color to show current years contribution as positive or negative
ggplot(account, aes(year, carryforwardnet, fill=net, group=year))+geom_col()+scale_fill_gradient(low="red", high="black")


# another example - find the maximum speed from a set of speeds - lets make up an example
# by sampling from a random uniform distribution
speeds = runif(min=0, max=100, n=300)

maxspeed=0
for ( i  in 1:length(speeds)) {
  maxspeed = ifelse(speeds[i] > maxspeed, speeds[i], maxspeed)
}

maxspeed
max(speeds)


```


Try this - make a random sample of fertilizer application, 10 values
with mean of *5* and standard deviation of *0.5*

(user *rnorm* function in R to do this)


Create a random sample of mean annual growth of a crop due to climate variation, 10 values with mean  of *20*, standard deviation of *2*,

Imagine that growth due to combined effects of climate and fertilize effects can be estimates as:

$ 1.8*fertilizer^2-0.5*fertilizer + mean annual growth $

Use a *for* loop to estimate combined growth for each of 10 samples

(you can assume that samples are matched - so you will have 10 growth estimates - using your 10 fertilizer applications and 10 estimates of mean annual growth)

Can you put together in a data structure that lets you plot growth against fertilizer?


# Do this on your own
# Review looping.Rmd - its a long one
