---
title: "Control Structures"
author: "Naomi Tague"
date: "January, 2022"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)

library(tidyverse)
library(purrr)
```

# Exchange functions

# Review DataTypes.Rmd

How do you figure out the rarest fish in our simulated ocean

Two key "take homes"

* working with *factors*
* how to return multiple items from a function using *list*

# Answer
```{r answer}

possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch1 = base::sample(possible.fish, size=20, replace=T)
rarestfish = names(which.min(summary(as.factor(catch1))))
rarestfish

plottitle = sprintf("The rarest fist is %s", rarestfish)
ggplot(data.frame(catch1=catch1), aes(catch1, fill=catch1))+geom_histogram(stat="count")+labs(title=plottitle)

```



# Flow Control (think of steering your program)

Another KEY concept is flow control
In your function you do different things depending on a conditions

CLASSIC example is 

IF then ELSE

If you have multiple conditions we can use case_when
Here's a silly simple example of how it works

# Simple example of flow control with if
```{r}

mycortest  = function(x,y, thresh=0.8) {
  # compute correlation
  res = cor(x,y)
  classification = ifelse(res > thresh, "GOOD", "NotGood")
  return(classification)
}

a = runif(min=1, max=100, n=100)
b = runif(min=1, max=100, n=100)

mycortest(a,b)
mycortest(a,a)
# this doesn't work - why?
mycortest(a, 1)

# useful to add error checking
mycortest  = function(x,y, thresh=0.8) {
  # compute correlation
  if(length(x) != length(y)) {return("unequal lengths for x and y")}
  res = cor(x,y)
  classification = ifelse(res > thresh, "GOOD", "NotGood")
  return(classification)
}
mycortest(a, 1)

```

# Flow control with a simple if
```{r fc}

# Simple "IF*
# imagine we are trying to get a tuna  - lets "fish" by sampling
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch1 = base::sample(possible.fish, size=1, replace=T)
catch1
ifelse(catch1 == "tuna", "success", "tryagain")

catch1 = "tuna"
ifelse(catch1 == "tuna", "success", "tryagain")
```

# multiple alternatives 

 what if we have more than one category of fish - grade A, B, C
steelhead are A, tuna are B and everything else is C

R and other languages have ways to do this multiple alternatives flow control
in R an example is

*case_when*

case_when(

condition ~ response,

condition ~ response
...
)

```{r was}

fish = "steelhead"

case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")

# apply to all of our ocean
# start by making a function
classify_fish = function(fish) {
 class=  case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
 return(class) }

classify_fish("steelhead")
classify_fish("cod")

# apply to our ocean
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch2 = base::sample(possible.fish, size=20, replace=T)
catch2

classify_fish(catch2)

```

# A more interesting example

Lets imagine that we are monitoring pollution in a lake, and we want to write a function that will let us know (flag) if risk associated with nutrient pollution are high, medium or low

From ecological studies, we know that
Risk is high if water temperature is greater than a threshold for more than 5 days, 
and mean nutrient concentration is greater than a high threshold

Risk is medium if water temperature is greater than a  threshold for more than 5 days
and mean nutrient concentration is greater than a medium threshold



Inputs:   
nutrient concentration for at least 5 days
temperature for at least 5 days
thresholds for temperature and nutrient (with default values)

Output:

Mean Nutrient Concentration
Pollution Risk as "low", "med" or "high"

Take a look at classify_lake 
* example of while
* example of flow control with if..else   and case_when

```{r}

source("R/classify_lake.R")

# generate some data to try the function
temperature = runif(min=4, max=35, n=30)
nutrient = runif(min=5, max=40, n=30)

# try it
res=classify_lake(temperature=temperature, nutrient=nutrient)

# lets pick some numbers we know should give us a low value
nutrient = runif(min=0, max=5, n=30)
classify_lake(temperature=temperature, nutrient=nutrient)

# now try high
# use repeat to generate the same value multiple times
nutrient = rep(50, times=30)
temperature = rep(25, times=30)
classify_lake(temperature=temperature, nutrient=nutrient)
```



# What we've learned

* how to write a function (and add error checking)
* how to generate data
* how to repeat in code  (different types of looping)
* how to make choices (flow control)



# Assignment -on your own

Write a function that takes a vector of fish names and always returns three items

* the most common fish, 
* the rarest fish
* the total number of fish

Create an Rmarkdown to demonstrate the use of your function with fish.txt - which
is under Data on ESM_262_Examples

[Data on ESM_262_Examples]{https://github.com/naomitague/ESM_262_Examples/blob/main/Data/fish.txt}


Turn in on Gauchospace what your function returns when you run with fish.txt!

Challenge: What if we had multiple catches - how would you run your summary function for all of those catches - see below for an example to generate multiple catches
You don't have to run this one in but we will go over in class

```{r  challenge}
# generate some data if we had 10 different fishing days
# some examples of using the purr family of functions

possible.fish = c("salmon","steelhead","shark","tuna","cod")

# lets generate random sample of the number of fish caught on each day
ndays=10
catches = round(runif(min=3, max=400, n=ndays))

# now for each fishing day (and its number of fish caught) - use sample to genreate the fish
catches_res = list(catches) %>% pmap(sample, x=possible.fish, replace=TRUE)
# look at the sample catches
catches_res[[1]]
catches_res[[10]]

# ok so how do we get the summary (rarest fish, number of fish for all catches in catch_res)?

```


Also review the classify_lake.R function from today's lecture to make sure you understand
