---
title: 'Looping'
author: "Naomi Tague"
date: "January, 2023"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```
# Comments/Questions 

looping.RMD
# Looping (or repetition) 

Looping - or repeating applying your function over multiple inputs is not always automatic 
  
  * more complex scenarios - number of different values is not the same for all inputs
    (e.g - you have 3 different speeds for 5 different cars (mass/Area))
  
  * other programming languages

Lets start with the basics of how you do repetition in programming in general often called *looping*

# For loops

 **for** statement 
  
* defines a loop using a counter that is incremented each time you go through the loop
  
* the counter - could be any variable (we often use *i*) but you could
  use what ever variable you want
  
* the loop is  the commands between the *{* and *}* after the *for* keyword
  
* these commands are repeated each time you go through the loop (following the definition)

```{r loop }

# simple example 
# a = 0+1+2+3+4+5
a=0
for (i in 1:5) {
 a = a+i
}
a

#if you want to keep track of a for each iteration
# start with a data structure to hold the results - there will be 5 iterations

a = rep(x=0, times=5)
for (i in 1:5) {
 a[i] = a[i]+i
}
a

# lets say we wanted to look income minus expenses for net income
# create some data for an example

income = runif(min=1000, max=5000, n=10)
income
expenses = rnorm(mean=1500, sd=500, n=10)
expenses
# it is possible that this approach could generate negative expenses
expenses = ifelse(expenses < 0, 0, expenses)
# normally we can just do
net = income-expenses
net


# lets make this a data frame for nice plotting
account = data.frame(income=income, expenses=expenses, net = net)
ggplot(account, aes(income, expenses))+geom_point()

# lets add a year
account$year = seq(from=2000, length.out=nrow(account))


# we could have computed net income with a for loop
# sometimes we use NA to show we haven't computed it yet, so "initialize" as NA
netloop = rep(NA, times=10)
# alternatively we could add to account
account$netloop=NA
account
# note I can use any variable I want to as a counter - 
# notice how I'm now using columns of the data frame
for (n in 1:10) {
  account$netloop[n] = account$income[n]-account$expenses[n]
}

# as expected net and netloop are the same :)
head(account)
```

# Why loop

what if we wanted to combine different rows and different columns?

for example, we wanted to carry forward from the previous year (info from a different row)

so carry forward net income is current years net income plus last years net

```{r loop2 }

#View(account)

# note: we treat first year differently because we don't have carry forward
# initialize our column to zero, start counter (n) at 2 instead of 1
# make the first value equal to the net of first year

account$carryforwardnet = 0
account$carryforwardnet[1]=account$net[1]
# now loop - looking backward one year to get previous years carry forward
for (n in 2:10) {
  account$carryforwardnet[n] = account$net[n]+account$carryforwardnet[n-1]
}

# plot - and use color to show current years contribution as positive or negative
ggplot(account, aes(year, carryforwardnet, fill=net, group=year))+geom_col()+scale_fill_gradient(low="red", high="black")


# another example - find the maximum speed from a set of speeds - lets make up an example
# by sampling from a random uniform distribution
speeds = runif(min=0, max=100, n=300)

maxspeed=0
for ( i  in 1:length(speeds)) {
  maxspeed = ifelse(speeds[i] > maxspeed, speeds[i], maxspeed)
}

maxspeed
max(speeds)


```

# Another Example
Try this - make a random sample of fertilizer application, 10 values
with mean of *5* and standard deviation of *0.5*

(user *rnorm* function in R to do this)

<span style="color:orange">Function to computed crop yield</span>

Lets imagine that annual yield of a crop can be estimated follows:


$$ yield =  1.8*fertilizer^2 - 0.5*fertilizer  + 0.1*TP $$


and TP is mean precipitation in cm


<span style="color:gold">Task 1</span>: Use a *for* loop to compute the total yield after 10 years

  * fertilizer is normally distributed with mean of 5 and standard deviation of 0.5
  * TP is 20cm
  
  
<span style="color:gold">Task 2</span>: create a function to that takes as input a single value for annual fertilizer and annual total precipitation and returns yield


<span style="color:gold">Task 3</span>: modify your function so that it returns total yield if the user inputs a vector of fertilizer and a vector of precipitation

Check that function returns the same result as your original *for* loop

Add some error checking and test

# Do on your own first before you check with this

```{r cropyield}

fert = rnorm(n=10, mean=5, sd=0.5)
TP = 20
crop = data.frame(fert=fert)
crop$yield = NA
for (yr in 1:10) {
  crop$yield[yr] = 1.8*fert[yr]^2 - 0.5*fert[yr]  + 0.1*TP
}

sum(crop$yield)
ggplot(crop, aes(fert, yield))+geom_point()+labs(y="yield", x="fertilizer")

# as a function

compute_yield = function(fert, TP) {
  yield = 1.8*fert^2-0.5* fert + 0.1*TP
  return(yield)
}

# apply function
compute_yield(fert, TP)
sum(compute_yield(fert, TP))

# a different function that includes annual sum
compute_total_yield = function(fert, TP) {
  yield = 1.8*fert^2-0.5* fert + 0.1*TP
  ty = sum(yield)
  return(ty)
}

compute_total_yield(fert, TP)

# useful error checking
compute_total_yield = function(fert, TP) {
  
  fert = ifelse(fert < 0, return("fertilizer cannot be negative"), fert)
   
  
  yield = 1.8*fert^2-0.5* fert + 0.1*TP
  ty = sum(yield)
  return(ty)
}
```




# Steps for running function over multiple inputs 


1. design a data structure to store results: sometimes this is automatic but not always
2. generate the  data (or read it in)
3. find ways to repeat things efficiently
  * *purr* family of function in R
  * more generally looping (*for*, *while*, ...others)
  

#  <span style="color:blue"> A bit more on Looping 

Loops can be "nested" with one loop inside the other

"Nesting" means loops inside loops

useful for "doing" something to each row and each column


**FOR** (i in 1:nrow) {

  **FOR** (j in 1: ncol)
  
  ...
  
}}


In nested loops, the inner loop (last **FOR** in the sequences, gets done first)

So the loop above will all columns for the first row and then go to the next row


# Example

Calculate NPV for

* a range of different interest rates and 
* a range of damages 
* that may be incurred 10 years in the future

**Steps**

* define inputs (interest rates, damages)
* define output (NPV)
* write the function
* create a data structure to store results where we vary both interest rates and damages
* use nested for loops to fill in the data structure

Try it first...

# Rcode
```{r npvfor, echo=TRUE, error=FALSE}

# write a function to compute npv
source("../R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


#generate some input
damages = c(25,33,91,24)
damages
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates
yr=10

# note that the "simple" application doesn't work - why?
compute_NPV(20, discount=discount_rates, value=damages)

# compute some npv's for different discount rates for each damage
# first generate a dataframe to store results - we have both damages and discount rates that are changing - so we need rows and columns
npvs = data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# here we have a 2-dimensional array - rows and columns
View(npvs)

# now use a nested for loop to populate
 for (i in 1:length(damages)) {
         for (j in 1:length(discount_rates)) {
       npvs[i,j]= compute_NPV(value=damages[i],       discount=discount_rates[j],time=yr )

         }
 }
 npvs
 
 
 #some data wrangling to make it pretty
colnames(npvs)=discount_rates
rownames(npvs)=damages
 npvs
 
# how do I plot this with ggplot - add a column for original value and then rearrange
 npvs$damage = damages
 npvsg =npvs %>% pivot_longer(!damage,names_to="dis", values_to="npv")
 head(npvsg)
 ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")
 
 # how about summing all the damages
 npv.total =npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
 ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

  
```

# An alternative method

An alternative (R specific) is to use functions from the **purr** package

Specifically *pmap* - which applies a function over multiple inputs 

```{r}
 # we could also do the repetition with map_dfc from the purr family of functions

# use damages and discount rates from above

discount_rates
damages

# with purr family, pmap runs the function for each value and returns as a data frame 

npvs = pmap(list(discount_rates),compute_NPV, time=10, value=damages)
# turn into a data frame
npvs = as.data.frame(npvs, col.names=discount_rates)
colnames(npvs) = discount_rates
npvs$damage = damages
# clean up and plot
 npvsg =npvs %>% pivot_longer(!damage,names_to="dis", values_to="npv")
 head(npvsg)
 ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")
 
 
 # how about summing all the damages
 npv.total =npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
 ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

 
 
```
# Some other types of loops

* while
  useful for repeating until a condition is met

Example
if a metal toxin in a lake increases by 1% per year, how many years will it take for the metal level to be greater than 30 units, if toxin is current at 5 units


```{r} 

# accumulate pollutant until a threshold - how many years does it take

# initial conditions
yr=1
pollutant_level = 5

# loop
while ((pollutant_level < 30) &&(yr < 1000))  {
  # increase pollutant
pollutant_level = pollutant_level + 0.01* pollutant_level 
# keep track of time
yr = yr + 1
}

pollutant_level
yr

# while loop dangers
# what if it doesn't end....

# see below - I left it commented out -because it will run for ever :)
# you can use controlC to stop

#while ((pollutant_level < 30))  {
  # increase pollutant
#pollutant_level = pollutant_level - 0.01* pollutant_level 
#}


```

# A bit more on nesting loops

we could even have multiple nesting (think about doings something to a 3 dimensional matrix 
time and space (x,y)

useful for "doings" something to each row and each column

    **FOR** (x in 1:nrow) {

        **FOR** (y in 1: ncol)  
  
          **FOR**  (t in 1: ntime)
 
 
                ...
  
        }
      }
    }



To illustrate
```{r}

# lets ocean be a 3 dimensional array - space (x,y) and time
# lets say is a 2 by 3 spatial maps (2 units across, 3 units down), and 2 time steps
ocean = array(dim=c(2,3,2))

ocean

# lets make up data for our ocean

ocean[1,1,1] = runif(n=1, min=0, max=1)

# second time step
ocean[1,1,2] = runif(n=1, min=0, max=1)

# next column, first time step
ocean[2,1,1] = runif(n=1, min=0, max=1)

# for loops let us do this quickly
for (x in 1:nrow(ocean)) {
  for (y in 1:ncol(ocean)) {
    for (t in 1:dim(ocean)[3]) {
      ocean[x,y, t] = runif(n=1, min=0, max=1)
    }}}


```


# Assignment 3

For this assignment you will write a new Rmarkdown document based on the analysis of power required for different *possible_cars* from looping.RMD; You will

* Add an additional super light car with mass 5,000 kg and a surface area 10m2

* Compute the mean and maximum power for Cars A, B, C and for your new car, assuming that average highway speed is 80 km/hr with a standard deviation of 10km/hr;  Use 2 different methods to compute the mean and maximum power for each car type; a) **FOR** loop and b) the **pmap** function from **purr**

* Create two boxplots (one for each method (**FOR** and **pmap**)) that show the range of power consumption (across sampled speeds) for each car type.

Put the Rmarkdown in your assignment github and submit a link on Gauchospace for Assignment 3 when its ready to be graded

**Grading Rubric**

1. Correct R code for adding the super light car (5pts)
2. Values for mean and maximum power for all 4 cars are in expected ranges (remember you are sampling so values will differ slightly) (5pts)
3. Used both methods (**FOR** and **pmap**) correctly (4pts)
4. Boxplots (6pts)
  a) correctly show range of power consumption for each car type for the two methdos
  b) good presentation (e.g axis labels, colors)
5. Good coding practice (e.g Adequate documentation throughout) (5pts)

Total out of 25pts





# To Review for next class

* DataTypes.Rmd
