---
title: 'Testing'
author: "Naomi Tague"
date: "January, 2025"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

``` {r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
#knitr::opts_knit$set(root.dir = '/Users/christina/Desktop/Courses.local/ESM262n/docs/packages')
library(tidyverse)
library(devtools)
library(roxygen2)
library(testthat)


```

# Datatypes

* working with *factors*
* how to return multiple items from a function using *list*s

# <span style="color:orange"> Types of Testing

Many types ...

# <span style="color:orange">Unit Testing</span>

**Unit testing**

* test each unit
    * function
    * box in your flow chart  

* mostly focused on does each piece work!


Top 12 Reasons to Write Unit Tests - Burke and Coyner (Java programmers)

* Tests reduce bugs in new features
* Tests reduce bugs in existing features
* Tests defend agains other programmers
* Testing forces you to slow down and think
* Testing makes development faster
* Tests reduce fear

Model developers now often use software to help them automate the testing process
Re-uses tests - makes it efficient to repeat many tests as you develop and modify the code 

Particularly helpful when you have multiple modules (as in our mangrove example)

This type of software is available for R, Python, C etc.

# <span style="color:orange">Testing in R</span>

In R, *testthat* library 

Install and add this library


* need *devtools* and *testthat* libraries

* test_dir(“name”) :runs all tests in the “name” subdirectory  (all files beginning with the word “test”

* test_file(“name”): runs all  the tests in a file called “name”

What is a test?

Testing in R uses expectation

# <span style="color:orange">Expectation</span>

Functions that you can use to make sure your code is working the way you think it should be working

* expect_true(condition)
* expect_equal(value, value)
* expect_match(string1, string2)
* expect_null(value)
* expect_length(object, n)


A good source for more expectation code and ways to do testing

[testing](https://testthat.r-lib.org/reference/index.html)

```{r testingsimple, eval=TRUE, error=TRUE}


# runs a test on a statement and checks to see if it is correct
# this one should work
expect_equal(2+2, 4)
# this one should 'throw and error'
expect_equal(2+5, 4)

# we can also use expect_true
# works with variables two
a = 200
expect_true(a > 2 )
a = 0
expect_true(a > 2 )

# matches
fish1 = "salmon"
fish2 = "trout"
expect_match(fish1, fish2)

# length
a = seq(from=1, to=10)
expect_length(a, 10)
expect_length(a, 11)

```

# <span style="color:orange">Expectation</span>

We can use expectation with functions 

Quickly and repeatedly test that they are working as expected 
as you develop code

# <span style="color:orange">Try it</span>

Source *compute_NPV.R*

use *expect_true* to test whether this function works

What might be good tests?

* positive values
* zero conditions (no money)

```{r testfunctions}

source("../R/compute_NPV.R")

futurevalue = 100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)

futurevalue = -100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)


# what if function returns a list?
# we can use the built in summary function to see how this works
testdata = rnorm(mean=25, sd=3, n=100)
summary(testdata)
# it is actually returning a list, but what are the names?
names(summary(testdata))
# now we can specify them
summary(testdata)["Max."]
# and use in a test to make sure this function works -...hmm
# well the max should be greater than the min right?

expect_true((summary(testdata)["Max."]-summary(testdata)["Min."]) > 0)
```
# <span style="color:orange">Testing</span>

The *art* of testing is in creating the test

* known inputs - known outputs
* what happens if user gives you garbage data
* what inputs would test "functionality"

Examples?

# <span style="color:orange"> More on testing functions 

Lets create a function with an error 
 
We want a function that computes total rainfall and total recharge from a daily time series of rainfall
 
Daily recharge of water to soil after rain is:

$ recharge = daily rainfall - interception $

*inputs*: 

  * time series of daily  rainfall (mm), 
  * single value interception (default 1mm)

*outputs*: 

  * totalrainfall (mm), 
  * totalrecharge (mm)

What are some tests we should do?


```{r testfunctions2}

# Implement function
source("../R/compute_recharge.R")

# lets do some simple tests
# if it never rains for 730 days then both total rainfall and recharge  should be zero

# notice how we access the list items from multiple returns
raindata = rep(0, times=730)
expect_that(compute_recharge(raindata)$totalrain, equals(0))
# so far so good, now for recharge

raindata = rep(0, times=730)
expect_that(compute_recharge(raindata)$totalrecharge, equals(0))

# what went wrong???


```

Fix and retry

```{r}

source("../R/compute_recharge_corrected.R")
raindata = rep(0, times=730)
expect_that(compute_recharge(raindata)$totalrecharge, equals(0))
```


# <span style="color:orange">Testing in R</span>


A **test** runs a bunch of **expectation** functions and returns a message

You can also include data!


Format

test_that(description, series of expectations)

*Returns*

Information about how many expectations (tests) passed or failed

We usually store these in a subdirectory called **tests**

Example *test_compute_recharge.R*

```{r trytesting}
source("../R/compute_recharge.R")
test_file("../tests/test_compute_recharge.R")

# see what happens if we go back to old file
source("../R/compute_recharge_corrected.R")
test_file("../tests/test_compute_recharge.R")
```


# <span style="color:orange">Using tests </span>

What we really want is to have tests that we **always** run, everytime we change something


Store multiple test in a *name.R* file  store these in these in the *tests* directory

Two options

* To run all tests in a file, use *test_file*


* To run all tests in a directory,use *test_dir*  (you can also do this from Build menu)


*Goal*: Make tests for each function that might be part of a larger project

```{r testing, eval=FALSE}

source("../R/compute_NPV.R")
# put your test in a file
test_file("../tests/test_compute_NPV.R")


# run all tests
test_dir("../tests")
```

# <span style="color:orange">A bit more on testing</span>


It is good practice to create tests for functions (especially
useful if you are going to have multiple functions and/or multiple developers)

It make sure that the function returns what you think

Recall when we designed contracts at the beginning of class, testing is a way to make sure the program is honoring the contract

You need to be creative here - test for things you know will work (quick way to find bugs)

Usually we want to run multiple tests..


# <span style="color:orange"> Stepping back - What we've covered in this course

* Data Science starts with clearly defining your goal

* Turning goals into program
  * clearly defined pathway from inputs to output
  * modularity - breaking it in to steps
  * flow chart first - having a plan
  
* Core concepts
  * function
  * data types and structures
  * looping
  
* Best (and common!) Practices
  * git and git hub for collaborating and tracking
  * Documenting as you go 
  * Testing as you go

Thank you!

# Last Assignment! 

For this assignment work in *pairs* - and submit as a group

* Design a function - you can pick any subject and you can even make up the equations as long as they conceptually make sense
  * Make sure it has at least 2 inputs and 1 parameter (ideally more) and at least 2 outputs
  * Code your function in R, save as a R file in subdirectory called R
  * Make sure you include documentation (both at the top as we've shown in past examples, and inline)
  * Include some error checking

* In an Quatro file, generate some data for 2 of the function inputs
  * use a *for*  loop to run the function for the data
  * repeat the "looping" using something from the "purrr" package 
  * Graph results (you can decide what the most interesting way to graph - you just need to make one graph)

* Write at least 2 tests for your function; store in a separate test file

* Put this in a git repo that includes an R subdirectory, a tests subsdirectory and your Quatro file and submit the link on Canvas (you can put all of this in a new git repo or a subdirectory of an existing git repo)

Submit the link to repo on Canvas (and name of subdirectory if part of a larger git project) as a group with your partner,



*Grading Rubric*

* Created a function 
  * at least 2 inputs, 1 parameter and 2 outputs (it can have more) (5pts)
  * includes error checking (5 pts)

* Quatro file that generates test data and runs the function
  * data generation ( 5pts)
  * run function multiple times with a loop (5 pts)
  * run function multiple times with something from *purr* (pmap) (5pts)
  * Graphs results (5 pts)

* Test file
  * a file with two tests (10pts)

Good programming practices (10 pts)
Total 50pts

