---
title: 'Using functions with Data'
author: "Naomi Tague"
date: "January, 2022"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```

# Questions

* sampling_example.Rmd
* error_checking.Rmd

# Steps for running your function over multiple inputs

1. design a data structure to store results: sometimes this is automatic but not always
2. generate the input data
3. apply to the function



# Consider our function that estimates the power required to keep a car moving at a given speed
 Code

```{r powerexample}

#' Power Required by Speed
#'
#' This function determines the power required to keep a vehicle moving at 
#' a given speed
#' @param cdrag coefficient due to drag default=0.3 
#' @param crolling coefficient due to rolling/friction default=0.015
#' @param v vehicle speed (m/2)
#' @param m vehicle mass (kg)
#' @param A area of front of vehicle (m2)
#' @param g acceleration due to gravity (m/s) default=9.8
#' @param pair (kg/m3) default =1.2
#' @return power (W)

autopower = function(V, m, A, cdrag=0.3, crolling=0.015,pair=1.2,g=9.8) {
  P = crolling*m*g*V + 1/2*A*pair*cdrag*V**3
  return(P)
}

```

I've also stored the autopower.R function as a file = using the recommended naming convention

**autopower.R**


# Generating data for the function and iterating over that data

Two parts

* Generate a data structure to store your results

* Repeat application of your function over the data


Example use: Imagine we want to see how much power is generated given a scenario where we know the mean and standard deviation of vehicle speeds


```{r sampling}

source("../R/autopower.R")

# generate sample speeds from a distribution
nsample = 100
speeds = rnorm(mean=25, sd=4, n=nsample)

# Step 1  create data frame to store results 
# how many simulations, what do you want to keep

#create a dataframe that has rows for each model run
# columns for height, flowrate and power estimate
results = data.frame(speed=speeds, power=NA)

head(results)

# if you only have one input and everything else is the same, R is smart enough
# to create a set of outputs
results$power = autopower(V=speeds, A=25, m=20000)

# ALWAYS look at your results to make sure it make sense
ggplot(results, aes(speed, power/1000))+geom_point()+labs(x="speed in m/s", y="power in kW")
ggplot(results, aes(x="POWER", y=power/1000))+geom_boxplot()+labs(y="power in kW", x="")

```

# Looping (or repetition) 

Looping - or repeating applying your function over multiple inputs is not always automatic 
  
  * more complex scenarios - number of different values is not the same for all inputs
    (e.g - you have 3 different speeds for 5 different cars (mass/Area))
  
  * other programming languages

Lets start with the basics of how you do repetition in programming in general often called *looping*

# For loops

 **for** statement 
  
* defines a loop using a counter that is incremented each time you go through the loop
  
* the counter - could be any variable (we often use *i*) but you could
  use what ever variable you want
  
* the loop is  the commands between the *{* and *}* after the *for* keyword
  
* these commands are repeated each time you go through the loop (following the definition)

```{r loop }

# simple example 
# a = 0+1+2+3+4+5
a=0
for (i in 1:5) {
 a = a+i
}
a

#if you want to keep track of a for each iteration
# start with a data structure to hold the results - there will be 5 iterations

a = rep(x=0, times=5)
for (i in 1:5) {
 a[i] = a[i]+i
}
a

# lets say we wanted to look income minus expenses for net income
# create some data for an example

income = runif(min=1000, max=5000, n=10)
income
expenses = rnorm(mean=1500, sd=500, n=10)
expenses
# it is possible that this approach could generate negative expenses
expenses = ifelse(expenses < 0, 0, expenses)
# normally we can just do
net = income-expenses
net


# lets make this a data frame for nice plotting
account = data.frame(income=income, expenses=expenses, net = net)
ggplot(account, aes(income, expenses))+geom_point()

# lets add a year
account$year = seq(from=2000, length.out=nrow(account))


# we could have computed net income with a for loop
# sometimes we use NA to show we haven't computed it yet, so "initialize" as NA
netloop = rep(NA, times=10)
# alternatively we could add to account
account$netloop=NA
account
# note I can use any variable I want to as a counter - 
# notice how I'm now using columns of the data frame
for (n in 1:10) {
  account$netloop[n] = account$income[n]-account$expenses[n]
}

# as expected net and netloop are the same :)
head(account)
```

# Why loop

what if we wanted to combine different rows and different columns?

for example, we wanted to carry forward from the previous year (info from a different row)

so carry forward net income is current years net income plus last years net

```{r loop2 }

#View(account)

# note: we treat first year differently because we don't have carry forward
# initialize our column to zero, start counter (n) at 2 instead of 1
# make the first value equal to the net of first year

account$carryforwardnet = 0
account$carryforwardnet[1]=account$net[1]
# now loop - looking backward one year to get previous years carry forward
for (n in 2:10) {
  account$carryforwardnet[n] = account$net[n]+account$carryforwardnet[n-1]
}

# plot - and use color to show current years contribution as positive or negative
ggplot(account, aes(year, carryforwardnet, fill=net, group=year))+geom_col()+scale_fill_gradient(low="red", high="black")


# another example - find the maximum speed from a set of speeds - lets make up an example
# by sampling from a random uniform distribution
speeds = runif(min=0, max=100, n=300)

maxspeed=0
for ( i  in 1:length(speeds)) {
  maxspeed = ifelse(speeds[i] > maxspeed, speeds[i], maxspeed)
}

maxspeed
max(speeds)


```

# Another Example
Try this - make a random sample of fertilizer application, 10 values
with mean of *5* and standard deviation of *0.5*

(user *rnorm* function in R to do this)

<span style="color:orange">Function to computed crop yield</span>

Lets imagine that annual yield of a crop can be estimated follows:


$$ yield =  1.8*fertilizer^2 - 0.5*fertilizer  + 0.1*TP $$


and TP is mean precipitation in cm


<span style="color:gold">Task 1</span>: Use a *for* loop to compute the total yield after 10 years

  * fertilizer is normally distributed with mean of 5 and standard deviation of 0.5
  * TP is 20cm
  
  
<span style="color:gold">Task 2</span>: create a function to that takes as input a single value for annual fertilizer and annual total precipitation and returns yield


<span style="color:gold">Task 3</span>: modify your function so that it returns total yield if the user inputs a vector of fertilizer and a vector of precipitation

Check that function returns the same result as your original *for* loop

Add some error checking and test

# Do on your own first before you check with this

```{r cropyield}

fert = rnorm(n=10, mean=5, sd=0.5)
TP = 20
crop = data.frame(fert=fert)
crop$yield = NA
for (yr in 1:10) {
  crop$yield[yr] = 1.8*fert[yr]^2 - 0.5*fert[yr]  + 0.1*TP
}

sum(crop$yield)
ggplot(crop, aes(fert, yield))+geom_point()+labs(y="yield", x="fertilizer")

# as a function

compute_yield = function(fert, TP) {
  yield = 1.8*fert^2-0.5* fert + 0.1*TP
  return(yield)
}

# apply function
compute_yield(fert, TP)
sum(compute_yield(fert, TP))

# a different function that includes annual sum
compute_total_yield = function(fert, TP) {
  yield = 1.8*fert^2-0.5* fert + 0.1*TP
  ty = sum(yield)
  return(ty)
}

compute_total_yield(fert, TP)

# useful error checking
compute_total_yield = function(fert, TP) {
  
  fert = ifelse(fert < 0, return("fertilizer cannot be negative"), fert)
   
  
  yield = 1.8*fert^2-0.5* fert + 0.1*TP
  ty = sum(yield)
  return(ty)
}
```

# More formal error checking

R also has specified totols for returning error messages

You can also use *stop()* or *warning()* inside your function to 
alter the user

*stop()* stops the remainder of the functions from executing
*warning()* continues but lets the user know

```{r, errorchecking, error=TRUE}

source("../R/compute_total_yield.R")
compute_total_yield

compute_total_yield(fert, TP)

compute_total_yield(fert=c(-2, 2,2), TP=3)
compute_total_yield(fert=c(2,2,2), TP=0)
```




* Review looping.Rmd - its a long one

