---
title: "Control Structures"
author: "Naomi Tague"
date: "January, 2025"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)

library(tidyverse)
library(purrr)
```

# <span style="color:orange"> What we've learned

* how to write a function (and add error checking)
* how to generate data
* how to repeat in code  (different types of looping)


Today

* Data Types ( data you use in functions and loops)


# <span style="color:orange"> Flow Control 

Another KEY concept is flow control

Allowing your function to "do" different things depending on a conditions

CLASSIC example is 

*IF* then *ELSE*

If you have multiple conditions we can use *case_when*

Here's a silly simple example of how it works

#<span style="color:orange">  Fishing

Imagine you are recording fish catches at different locations
(we are going to simulate this by sampling )

Fish have grades...

grade A, B, C

* steelhead are A, 
* tuna are B 
* everything else is C

R and other languages have ways to do this multiple alternatives flow control
in R an example is

*case_when*

case_when(

condition ~ response,

condition ~ response
...
)

First a simple example where we only have one fist
```{r was}

fish = "steelhead"
case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
```


# apply to all of our ocean
# start by making a function
classify_fish = function(fish) {
 class=  case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
 return(class) }

classify_fish("steelhead")
classify_fish("cod")

# apply to our ocean
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch2 = base::sample(possible.fish, size=20, replace=T)
catch2

classify_fish(catch2)

```

# A more interesting example

Lets imagine that we are monitoring pollution in a lake, and we want to write a function that will let us know (flag) if risk associated with nutrient pollution are high, medium or low

From ecological studies, we know that

*High Risk Conditions*

* risk is high if water temperature is greater than a threshold for more than 5 days, 
and 
* mean nutrient concentration is greater than a high threshold

*Medium Risk Conditions*

* Risk is medium if water temperature is greater than a  threshold for more than 5 days
and
* mean nutrient concentration is greater than a medium threshold

*Eveything else is low risk*

# Design the function

**Inputs**:   

* nutrient concentration for at least 5 days
* temperature for at least 5 days
* thresholds for temperature and nutrient (with default values)

**Output**:

* Mean Nutrient Concentration
* Pollution Risk as "low", "med" or "high"


Take a look at classify_lake 

* example of while
* example of flow control with if..else   and case_when

```{r}

source("../R/classify_lake.R")

# generate some data to try the function
temperature = runif(min=4, max=35, n=30)
nutrient = runif(min=5, max=40, n=30)

# try it
res=classify_lake(temperature=temperature, nutrient=nutrient)

# lets pick some numbers we know should give us a low value
nutrient = runif(min=0, max=5, n=30)
classify_lake(temperature=temperature, nutrient=nutrient)

# now try high
# use repeat to generate the same value multiple times
nutrient = rep(50, times=30)
temperature = rep(25, times=30)
classify_lake(temperature=temperature, nutrient=nutrient)
```




