---
title: "Control Structures"
author: "Naomi Tague"
date: "January, 2023"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)

library(tidyverse)
library(purrr)
```

#  Questions from Looping.RMD

# Some other types of loops

* while
  useful for repeating until a condition is met

Example
if a metal toxin in a lake increases by 1% per year, how many years will it take for the metal level to be greater than 30 units, if toxin is current at 5 units


```{r} 

# accumulate pollutant until a threshold - how many years does it take

# initial conditions
yr=1
pollutant_level = 5

# loop
while ((pollutant_level < 30) &&(yr < 1000))  {
  # increase pollutant
pollutant_level = pollutant_level + 0.01* pollutant_level 
# keep track of time
yr = yr + 1
}

pollutant_level
yr

# while loop dangers
# what if it doesn't end....

# see below - I left it commented out -because it will run for ever :)
# you can use controlC to stop

#while ((pollutant_level < 30))  {
  # increase pollutant
#pollutant_level = pollutant_level - 0.01* pollutant_level 
#}


```


# Flow Control (think of steering your program)

Another KEY concept is flow control

Allowing your function to "do" different things depending on a conditions

CLASSIC example is 

*IF* then *ELSE*

If you have multiple conditions we can use *case_when*

Here's a silly simple example of how it works

# Simple example of flow control with if
```{r}

mycortest  = function(x,y, thresh=0.8) {
  # compute correlation
  res = cor(x,y)
  classification = ifelse(res > thresh, "GOOD", "NotGood")
  return(classification)
}

a = runif(min=1, max=100, n=100)
b = runif(min=1, max=100, n=100)

mycortest(a,b)
mycortest(a,a)
# this doesn't work - why?
mycortest(a, 1)

# useful to add error checking
mycortest  = function(x,y, thresh=0.8) {
  # compute correlation
  if(length(x) != length(y)) {stop("unequal lengths for x and y")}
  res = cor(x,y)
  classification = ifelse(res > thresh, "GOOD", "NotGood")
  return(classification)
}
mycortest(a, 1)

```

# Flow control with a simple if
```{r fc}

# Simple "IF*
# imagine we are trying to get a tuna  - lets "fish" by sampling
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch1 = base::sample(possible.fish, size=1, replace=T)
catch1
ifelse(catch1 == "tuna", "success", "tryagain")

catch1 = "tuna"
ifelse(catch1 == "tuna", "success", "tryagain")
```

# Multiple Alternatives 

What if we have more than one category of fish 

grade A, B, C

* steelhead are A, 
* tuna are B 
* everything else is C


R and other languages have ways to do this multiple alternatives flow control
in R an example is

*case_when*

case_when(

condition ~ response,

condition ~ response
...
)

```{r was}

fish = "steelhead"

case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")

# apply to all of our ocean
# start by making a function
classify_fish = function(fish) {
 class=  case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
 return(class) }

classify_fish("steelhead")
classify_fish("cod")

# apply to our ocean
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch2 = base::sample(possible.fish, size=20, replace=T)
catch2

classify_fish(catch2)

```

# A more interesting example

Lets imagine that we are monitoring pollution in a lake, and we want to write a function that will let us know (flag) if risk associated with nutrient pollution are high, medium or low

From ecological studies, we know that

*High Risk Conditions*

* risk is high if water temperature is greater than a threshold for more than 5 days, 
and 
* mean nutrient concentration is greater than a high threshold

*Medium Risk Conditions*

* Risk is medium if water temperature is greater than a  threshold for more than 5 days
and
* mean nutrient concentration is greater than a medium threshold

*Eveything else is low risk*

# Design the function

**Inputs**:   

* nutrient concentration for at least 5 days
* temperature for at least 5 days
* thresholds for temperature and nutrient (with default values)

**Output**:

* Mean Nutrient Concentration
* Pollution Risk as "low", "med" or "high"


Take a look at classify_lake 

* example of while
* example of flow control with if..else   and case_when

```{r}

source("../R/classify_lake.R")

# generate some data to try the function
temperature = runif(min=4, max=35, n=30)
nutrient = runif(min=5, max=40, n=30)

# try it
res=classify_lake(temperature=temperature, nutrient=nutrient)

# lets pick some numbers we know should give us a low value
nutrient = runif(min=0, max=5, n=30)
classify_lake(temperature=temperature, nutrient=nutrient)

# now try high
# use repeat to generate the same value multiple times
nutrient = rep(50, times=30)
temperature = rep(25, times=30)
classify_lake(temperature=temperature, nutrient=nutrient)
```



# What we've learned

* how to write a function (and add error checking)
* how to generate data
* how to repeat in code  (different types of looping)
* how to make choices (flow control)



# Assignment 4

For this assignment you will write a new Rmarkdown document based on the analysis of power required for different *possible_cars* from looping.RMD; You will

* Add an additional super light car with mass 5,000 kg and a surface area 10m2

* Compute the mean and maximum power for Cars A, B, C and for your new car, assuming that average highway speed is 80 km/hr with a standard deviation of 10km/hr;  Use 2 different methods to compute the mean and maximum power for each car type; a) **FOR** loop and b) the **pmap** function from **purr**

* Create two boxplots (one for each method (**FOR** and **pmap**)) that show the range of power consumption (across sampled speeds) for each car type.

Put the Rmarkdown in your assignment github and submit a link on Gauchospace for Assignment 3 when its ready to be graded

**Grading Rubric**

1. Correct R code for adding the super light car (5pts)
2. Values for mean and maximum power for all 4 cars are in expected ranges (remember you are sampling so values will differ slightly) (5pts)
3. Used both methods (**FOR** and **pmap**) correctly (4pts)
4. Boxplots (6pts)
  a) correctly show range of power consumption for each car type for the two methdos
  b) good presentation (e.g axis labels, colors)
5. Good coding practice (e.g Adequate documentation throughout) (5pts)

Total out of 25pts



