---
title: "git_setup"
output: "html"
date: "2025-01-24"
author: "Naomi Tague"
---

```{r setup, include=FALSE}
library(tidyverse)
library(purrr)

```

# Introduction to Setting Up Git for Use with RStudio

Version control is essential for managing changes to your work and collaborating. Git is the most popular version control system and yay! integrates seamlessly with RStudio

I'm going to give you some basic instructions to get started...want more details?
This is a great reference

[Happy Git](https://happygitwithr.com/)

## Setup

## Step 1 GitHub

*Github* is an online server that manages the version control system - by using *Github* you can access your code from anywhere. 

Usually, code development is done locally using *Git* (in our case *Git* integrated with Rstudio), but by coordinating with the centralized *Github* , you can keep track of changes from anywhere; and allow multiple to work on the code at the same time (each on their local computers) - Believe it or now it works :)

To access *Github* you need to set up an account

Step 1: Set up your account (remember your username)

[Link to github](https://github.com)

# Step 2: Installing *git* on your computer (for use with Rstudio)

Happy git 
does a really nice job of taking you through this, so I'm going to refer you to that chapter for some step by step instructions

[Happy Git](https://happygitwithr.com/)

Note if you don't know what a shell (or Terminal window in R) is also look at Appendix A.

[Install GIT](https://happygitwithr.com/install-git)

**Verify Git Installation**: To check if Git is installed, open RStudio and go to `Tools > Shell...`. In the shell window that opens, type:

   ```
   git --version
   ```
Press Enter. If Git is installed, you should see a version number (e.g., `git version 2.x.x`). If not, check to make sure you've completed the process

Ask for help if you can't make this work

# Tell R about Git 


1. Open RStudio and go to `Tools > Global Options > Git/SVN`.
2. Check the box labeled **Enable version control interface for RStudio projects**.
3. If Git is not automatically detected, specify the path to the Git executable:
   - On Windows: The path is usually `C:\Program Files\Git\bin\git.exe` or similar.
   - On macOS/Linux: Use the Finder or File Explorer to locate where Git is installed, or ask for assistance.
4. Click **Apply** or **OK** to save your settings.

Let your local *Git* know who you are (GitHub user name and password)
```{r eval=FALSE}
library(usethis)
# substitute your user name before running
use_git_config(user.name = "Jane Doe", user.email = "jane@example.org")
```

## Step 3: Create or Link a Git Repository in RStudio

### Option 1: Create a New Project with Git

1. Go to `File > New Project > New Directory > New Project`.
2. Check the box labeled **Create a Git repository**.
3. Complete the setup by selecting a folder and naming your project.
4. Click **Create Project**. RStudio will initialize a Git repository for the project.

### Option 2: Link an Existing Project to Git

1. Open an existing project in RStudio.
2. Go to `Tools > Project Options > Git/SVN`.
3. Select **Git** as the version control system and click **OK**. This will initialize Git for the project.

Once Git is enabled, the **Git** tab will appear in the top-right pane of RStudio

Remember this Project, add your Automobile power function and Rmarkdown to it
We will cover next steps in class


## Troubleshooting 

- If the **Git** tab doesn’t appear, check that Git is enabled under `Global Options`.
- Ensure that Git is installed and correctly configured in RStudio.
- If you encounter issues, try restarting RStudio or reinstalling Git.

